---
title: "Take-Home assigment 3"
author: "Oliver, Alejo, Tirdod"
date: "2025-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, include = F}
library(sf)
library(spData)    
library(tidyverse)
library(dplyr)
library(gdistance)
library(terra)
library(exactextractr)
library(lubridate)
library(caret)

#devtools::install_github("statsbomb/StatsBombR")
#devtools::install_github("josephmisiti/SBpitch")
library(SBpitch)

library(tidyverse)
library(StatsBombR)

```

# Introduction

As football enthusiasts, we are keen on analyzing games in great depth. Fortunately, the growing availability of open football data provides many opportunities to do so. This was one of the main motivations for our project.
We set out to explore how shots have evolved over time, and whether we could build our own expected goals (xG) model — a tool widely used in football analytics to estimate the probability of a shot resulting in a goal.

To do this, we use open-access data provided by StatsBomb which includes detailed event-level information of football matches of different competition. It also contains StatsBomb’s own xG estimates. This allows us to compare our model’s predictions with this already established estimate.

Ultimately, our goal is to test the hypothesis that closer shots have a higher chance of resulting in goals, and to see whether this insight is reflected in trends over time. In other words: are teams becoming more conservative, favoring high-probability chances over long-range efforts?
In the end, we might see this being reflected in our xG model. The code for this project can also be found  on [GitHub](https://github.com/otausendschoen/Football_Analysis).




# Set up: Data & Libraries

For this project, we will rely on the package StasBomb. This package gives us access to detailed, event-level football data — including passes, shots, dribbles, and more — from professional matches around the world.

In general, the package is used in the following way:

- Load a set of competitions and matches
- Download all event data for those matches
- Do analysis on subsets of this data. For example, we will start analyzing how shots behave over time below.


In general, we can use the following code to obtain the data via the package's API. However, the dataset is very big so we decided to attach it as a CSV. Nevertheless, the code we used to obtain the initial data is below, commented out.


``` {r}

# #Pulling StatsBomb Free Data Into R
# library(tidyverse)
# library(StatsBombR)
# Comps <- FreeCompetitions()
# comps_shots <- Comps %>%
#   filter(
#     competition_gender == 'male',
#     !competition_name %in% c('FIFA U20 World Cup', 'Indian Super league', 'Major League Soccer', 'North American League')
#   )
# 
# Matches <- FreeMatches(Comps)
# 
# Matches_Shots <- Matches
# 
# Matches_Passes <- Matches %>%
#   filter(year(match_date) >= 2000)
# 
# StatsBombData_Shots <- free_allevents(MatchesDF = Matches_Shots, Parallel = T)
# StatsBombData_Passes <- free_allevents(MatchesDF = Matches_Passes, Parallel = T)
# 
# StatsBombData_Shots = allclean(StatsBombData_Shots)
# StatsBombData_Passes = allclean(StatsBombData_Passes)
# 
# shots <- StatsBombData_Shots %>%
#   filter(type.name == "Shot", !is.na(location)) %>%
#   unnest_wider(location, names_sep = "_") %>%
#   rename(x = location_1, y = location_2)
# 
# passes <- StatsBombData_Passes %>%
#   filter(type.name == "Pass", !is.na(location)) %>%
#   unnest_wider(location, names_sep = "_") %>%
#   rename(x = location_1, y = location_2)
# 
# shots <- shots %>%
#   left_join(
#     Matches %>%
#       select(match_id, match_date),
#     by = "match_id"
#   ) %>%
#   left_join(
#     Comps %>%
#       select(competition_id, season_id, competition_name, season_name),
#     by = c("competition_id", "season_id")
#   ) %>%
#   mutate(match_date = as.Date(match_date))
# 
# shots_clean <- shots %>%
#   select(
#     -carry.end_location,
#     -goalkeeper.end_location,
#     -tactics.lineup,
#     -related_events,
#     -shot.freeze_frame,
#     -pass.end_location
#   ) %>%
#   unnest_wider(shot.end_location, names_sep = "_") %>%
#   rename(
#     shot.end_x = shot.end_location_1,
#     shot.end_y = shot.end_location_2
#   )
# 
# passes <- passes %>%
#   left_join(
#     Matches %>%
#       select(match_id, match_date),
#     by = "match_id"
#   ) %>%
#   left_join(
#     Comps %>%
#       select(competition_id, season_id, competition_name, season_name),
#     by = c("competition_id", "season_id")
#   ) %>%
#   mutate(match_date = as.Date(match_date))
# 
# passes_clean <- passes %>%
#   select(
#     -carry.end_location,
#     -goalkeeper.end_location,
#     -tactics.lineup,
#     -related_events,
#     -shot.end_location,
#     -shot.freeze_frame
#   ) %>%
#   unnest_wider(pass.end_location, names_sep = "_") %>%
#   rename(
#     pass.end_x = pass.end_location_1,
#     pass.end_y = pass.end_location_2
#   )
# 
# write_csv(shots_clean, "shots.csv")
# write_csv(passes_clean, "passes.csv")

```

Another important point is that the dataset for shots is already filtered for the specific competition we use. This was the only possibility to get data as it was too big otherwise. On the other hand, the data for passes only includes 2 specific seasons for computational efficiency - and this is already includes about 200,000 observations.

```{r}
shots_df <- read.csv("shots.csv")
passes_df <- read_csv(unz("passes.csv.zip", "passes.csv"))
```

Before addressing our research question, we shall do some some preliminary data exploration to understand and get an overview of the data.

```{r}
head(shots_df[, -c(1,2,3)])
```

As we can see, there are in total 186 columns and 12720 observations. These observations are all seasons for most tournaments but filtered for only shots. The observations for passes on the other hand can be found in the other dataset.

Before we start the analysis, we do some basic data pre processing and also remove some of these redundant 186 columns that we do not require.

## Data pre-processing

```{r}


shots <- shots_df %>%
  dplyr::select(
    id,
    match_id,
    competition_name,
    season_name,
    timestamp,
    minute,
    second,
    period,
    team.name,
    possession_team.name,
    player.id,
    player.name,
    #position.id,
    position.name,
    #play_pattern.id,
    play_pattern.name,
    under_pressure,
    location.x,
    location.y,
    shot.end_x,
    shot.end_y,
    shot.end_location.z,
    shot.end_location_3,
    shot_impact_height,
    shot.statsbomb_xg,
    #shot.outcome.id,
    shot.outcome.name,
    #shot.technique.id,
    shot.technique.name,
    #shot.body_part.id,
    shot.body_part.name,
    #shot.type.id,
    shot.type.name,
    shot.aerial_won,
    shot.redirect,
    shot.follows_dribble,
    shot.first_time,
    shot.open_goal,
    shot.deflected,
    shot.redirect,
    shot.key_pass_id,
    player.id.GK,
    player.name.GK,
    shot.saved_off_target,
    shot.saved_to_post,
    location.x.GK,
    location.y.GK,
    AngleToGoal,
    AngleToKeeper,
    AngleDeviation
  )

passes<- passes_df %>%
  dplyr::select(
        id,
    match_id,
    competition_name,
    season_name,
    timestamp,
    minute,
    second,
    period,
    possession_team.name,
    team.name,
    player.id,
    player.name,
    position.name,
    play_pattern.name,
    under_pressure,
    location.x,
    location.y,
    pass.length, 
    pass.angle, 
    pass.end_x, 
    pass.end_y, 
    pass.aerial_won, 
    pass.switch, 
    pass.cross, 
    pass.assisted_shot_id, 
    pass.shot_assist, 
    pass.inswinging, 
    pass.deflected, 
    pass.outswinging, 
    pass.through_ball, 
    pass.cut_back, 
    pass.goal_assist, 
    pass.recipient.id, 
    pass.recipient.name, 
    pass.height.name, 
    pass.body_part.name, 
    pass.type.name, 
    pass.outcome.name, 
    pass.technique.name, 
    ball_receipt.outcome.name,
    pass.no_touch
  )

```


To make life easier in the next sections, we calculate the shot distance from player to goal.
StatsBomb uses a standard pitch coordinate system where:

- The field is 120 units long (from goal to goal, along the x-axis).
- The field is 80 units wide (from sideline to sideline, along the y-axis).
- The center of the opponent’s goal is at (120, 40): 
- 120 StatsBomb units ≈ 105 meters → 1 unit ≈ 0.875 meters

```{r}
calculate_shot_distance <- function(x, y) {
  sqrt((120 - x)^2 + (40 - y)^2) * (105 / 120)}

shots <- shots %>%
  mutate(shot_distance = calculate_shot_distance(location.x, location.y))

```

We do the same for the distance for the distance of the passes, from player to player.

```{r}
calculate_pass_distance <- function(x1, y1, x2, y2) {
  sqrt((x2 - x1)^2 + (y2 - y1)^2) * (105 / 120)
}

passes <- passes %>%
  mutate(pass_distance = calculate_pass_distance(location.x, location.y, pass.end_x, pass.end_y))

```

The shot.key_pass_id links the unique play id for a shot to the unique play id for the pass leading to the shot (if the shot was assisted). We do not need the unique play id for our purposes, but it will be helpful to know whether a shot was assisted or self-generated. We'll convert it to a true-false column.

``` {r}

shots <- shots %>%
  mutate(shot.is_assisted = !is.na(shot.key_pass_id)) %>%  # Create new column based on whether shot.key_pass_id is not null
  relocate(shot.is_assisted, .after = shot.key_pass_id) %>%  # Move is_assisted right after shot.key_pass_id
  dplyr::select(-shot.key_pass_id)    

```

# Data exploration

```{r}
shots %>%
  group_by(shot.outcome.name) %>%
  summarise(avg_distance = mean(shot_distance, na.rm = TRUE),
            count = n()) %>%
  arrange(avg_distance)

```

As we can see above, the average distance when scoring a goal is much lower compared to the other shot outcomes such as being blocked or being saved.

```{r}
passes %>%
  mutate(pass.outcome.name = replace_na(pass.outcome.name, "Complete")) %>%
  group_by(pass.outcome.name) %>%
  summarise(
    avg_distance = mean(pass_distance, na.rm = TRUE),
    count = n()
  ) %>%
  arrange(avg_distance)

```

Similarly, the average distance for completed passes is much, much shorter, than passes that are not completed. Our documentation from StatsBomb states that all completed passes have a null outcome name, so we input the completed pass value in this step as well.


```{r}

hist(shots$shot_distance, breaks = 30, main = "Shot Distance Distribution", xlab = "Distance to Goal (meters)")

```

``` {r}
shots %>%
  count(player.name, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(player.name, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Players by Total Shot Volume", x = "Player", y = "Shots")

```

Most importantly, the histogram shows how far most shots are taken from the goal. We observe a clear peak at shorter distances, indicating that many shots come from inside or near the penalty box. However, there are also a number of longer-range attempts, suggesting variation in shooting strategy across players or teams.

We can do the same for the pass distances.
```{r}
hist(passes$pass_distance, breaks = 50,
     main = "Distribution of Pass Distances",
     xlab = "Pass Distance (meters)",
     col = "lightblue", border = "white")
```

Let's plot this over time:

For convenience, we include a trendline using LOESS (Locally Estimated Scatterplot Smoothing), regression method that captures the underlying trend - without assuming a strictly linear relationship.

```{r}

avg_shot_distance_by_year <- shots %>%
  group_by(season_name) %>%
  summarise(
    avg_shot_distance = mean(shot_distance, na.rm = TRUE),
    shot_count = n()
  ) %>%
  filter(!is.na(season_name))

ggplot(avg_shot_distance_by_year, aes(x = season_name, y = avg_shot_distance, group = 1)) +
  geom_line(size = 1.2, color = "darkred") +
  geom_point(size = 2, color = "black") +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue", linetype = "dashed") +
  labs(
    title = "Average Shot Distance Over Time",
    x = "Season",
    y = "Average Distance to Goal (in pitch units)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

This plot shows how most shots are not within the penalty or 11m box from the goal line. Interestingly, we can see how the shot distance started to increase on average from 1970 on. However, the trend seems to have "stabilized at around 19 meter, indicating a potential tactical equilibrium — where players still attempt long-range shots, but the majority of efforts come from a more optimal shooting range. It will be interesting to see if we also observe this in our xG model.


The plot for the average distance of passes can also be seen below but it is not as meaningful as we only have data for 2 seasons. We can see a sharp decrease in distance.

```{r}
avg_pass_distance_by_year <- passes %>%
  group_by(season_name) %>%
  summarise(
    avg_pass_distance = mean(pass_distance, na.rm = TRUE),
    pass_count = n()
  ) %>%
  filter(!is.na(season_name))

ggplot(avg_pass_distance_by_year, aes(x = season_name, y = avg_pass_distance, group = 1)) +
  geom_line(size = 1.2, color = "darkgreen") +
  geom_point(size = 2, color = "black") +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue", linetype = "dashed") +
  labs(
    title = "Average Pass Distance Over Time",
    x = "Season",
    y = "Average Pass Distance (m)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


We will observe this more closely by taking a look at the distribution of each year. For this, we create a density plot faceted by season below.


```{r}
ggplot(shots, aes(x = shot_distance)) +
  geom_density(fill = "lightblue") +
  facet_wrap(~season_name) +  # no `scales = "free_y"`
  labs(
    title = "Distribution of Shot Distances per Season",
    x = "Shot Distance (m)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 6),
    strip.text = element_text(size = 8)
  )

```

However, the trend is more difficult to observe here. We can see how each season has a different pattern/shape and there is no universal rule or distribution that applies to all seasons.

```{r}
ggplot(passes, aes(x = pass_distance)) +
  geom_density(fill = "lightblue") +
  facet_wrap(~season_name) +
  labs(
    title = "Distribution of Pass Distances per Season",
    x = "Pass Distance (m)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 6),
    strip.text = element_text(size = 8)
  )

```

For passes on the other hand, it is way more obvious. We can definitely see that short passes are more likely in the 2023/2024 season than in the 2015/2016 season!


Next, we plot a few heatmaps in specific year to compare this in a bit more intuitive way over time.

# Heatmaps




# Expected Goals Model

Now, we'll create an expected goals model, to assess the likelihood of scoring a goal based on pre-shot information

``` {r}

xg_df <- shots %>%
  dplyr::select(shot.outcome.name,
         play_pattern.name,
         under_pressure,
         location.x,
         location.y,
         shot_distance,
         shot.technique.name,
         shot.body_part.name,
         shot.type.name,
         shot.aerial_won,
         shot.follows_dribble,
         shot.first_time,
         shot.open_goal,
         shot.is_assisted,
         location.x.GK,
         location.y.GK,
         AngleToGoal,
         AngleToKeeper,
         AngleDeviation
         ) %>%
  mutate(is_goal = shot.outcome.name == "Goal") %>%  # creates TRUE/FALSE
  dplyr::select(is_goal, everything(), -shot.outcome.name)  # moves is_goal to the first column

xg_df <- xg_df %>%
  rename(
    angle_to_goal = AngleToGoal,
    angle_to_keeper = AngleToKeeper,
    angle_deviation = AngleDeviation 
  ) %>%
  mutate(is_shot_foot = if_else(shot.body_part.name %in% c("Right Foot", "Left Foot"), TRUE, FALSE)) %>%
  relocate(is_shot_foot, .after = shot.body_part.name) %>%
  dplyr::select(-shot.body_part.name)

```

Next we perform categorical encoding on our columns that contain string values.
``` {r}

insert_one_hot <- function(df, col_name) {
  # Create dummy variables (as a data frame)
  dummies <- as.data.frame(model.matrix(~ . - 1, data = df[col_name]))
  
  # Clean up column names
  colnames(dummies) <- gsub(paste0("^", col_name), col_name, colnames(dummies))
  colnames(dummies) <- gsub(" ", ".", colnames(dummies))  # Replace spaces if needed
  
  # Get the original column position
  pos <- which(names(df) == col_name)
  
  # Build new df: before, dummies, after
  df_new <- bind_cols(
    df[1:(pos - 1)],
    dummies,
    df[(pos + 1):ncol(df)]
  )
  
  return(df_new)
}

xg_df_encoded <- xg_df  # make a copy to preserve original

cols_to_encode <- c("play_pattern.name", "shot.technique.name", "shot.type.name")

for (col in cols_to_encode) {
  xg_df_encoded <- insert_one_hot(xg_df_encoded, col)
}

xg_df_encoded <- xg_df_encoded %>%
  mutate(across(where(is.logical), ~ if_else(is.na(.), FALSE, .)))

```

Now, we're ready to start fitting our expected goals model.

Before we start by adding in some of our descriptive variables, we'll fit a simple univariate logistic regression to measure the impact of shot distance on goals scored
``` {r}

set.seed(42)  # for reproducibility

# Create an 70/30 train-test split based on the is_goal variable
train_index <- createDataPartition(xg_df_encoded$is_goal, p = 0.7, list = FALSE)
train_data <- xg_df_encoded[train_index, ]
test_data  <- xg_df_encoded[-train_index, ]

```

``` {r}

# Fit logistic regression on the training set
model_distance <- glm(is_goal ~ shot_distance, data = train_data, family = binomial)

# Summarize the model
summary(model_distance)

```

We can clearly see the convex curve represents that being closer to the goal has a direct impact on the probability of the goal.

Now, let's add the shot angle

``` {r}

# Fit logistic regression on the training set (now adding angle)
model_distance_and_angle <- glm(is_goal ~ shot_distance + angle_to_goal, data = train_data, family = binomial)

# Summarize the model
summary(model_distance_and_angle)

```

We evaluate the two models
``` {r}

AIC(model_distance, model_distance_and_angle)

anova(model_distance, model_distance_and_angle, test = "Chisq")

```

Adding shot angle on its own does not improve the model. With a p-value of .558, it is not statistically significant either.

Now, let's add whether the shot was under pressure
``` {r}

# Fit logistic regression on the training set (now adding whether the shot was under pressure)
model_distance_and_angle_and_pressure <- glm(is_goal ~ shot_distance + angle_to_goal + under_pressure, data = train_data, family = binomial)

# Summarize the model
summary(model_distance_and_angle_and_pressure)

```

We evaluate the three models
``` {r}

AIC(model_distance, model_distance_and_angle, model_distance_and_angle_and_pressure)

anova(model_distance, model_distance_and_angle, model_distance_and_angle_and_pressure, test = "Chisq")

```

We can see that the shot being under pressure has a sizable impact on goal probability. Meanwhile, the shot angle doesn't really have much of an impact, and it is not statistically significant either.

Now, let's try adding some more descriptive variables.
``` {r}

model_descriptive <- glm(is_goal ~ 
                           shot_distance + 
                           angle_to_goal + 
                           is_shot_foot + 
                           under_pressure +
                           shot.is_assisted
                    , data = train_data, family = binomial)

# Summarize the model
summary(model_descriptive)

```

``` {r}

AIC(model_distance_and_angle_and_pressure, model_descriptive)

anova(model_distance_and_angle_and_pressure, model_descriptive, test = "Chisq")

```

``` {r}

# Predict on the test set (get predicted probabilities)
predicted_probs <- predict(model_descriptive, newdata = test_data, type = "response")

# Create a prediction data frame covering the range of shot distances
pred_data <- data.frame(shot_distance = seq(min(train_data$shot_distance, na.rm = TRUE),
                                              max(train_data$shot_distance, na.rm = TRUE),
                                              length.out = 100))
pred_data$pred_prob <- predict(model_distance, newdata = pred_data, type = "response")

# Plot the relationship between shot distance and predicted probability
ggplot(pred_data, aes(x = shot_distance, y = pred_prob)) +
  geom_line(color = "blue") +
  labs(title = "Predicted Goal Probability vs. Shot Distance",
       x = "Shot Distance from the Goal (Meters)",
       y = "Predicted Probability of Scoring") +
  theme_minimal()

```


we can try using the expected goal predictions to compare that to actual goals scored. we'll use the 2015/16 German Bundesliga as an example
``` {r}



```