---
title: "FirstAnalysis"
author: "Alejandro Vacca"
date: "2025-03-21"
output: html_document
---

### Load libraries and Competitions using StatsBomb (FIFA Men's World Cup)

```{r}
library(tidyverse)  
library(StatsBombR)
library(SBpitch)
library(ggplot2)
library(dplyr)
library(StatsBombR)

# Get all available competitions
Comps <- FreeCompetitions()

# Filter only FIFA World Cup (male)
Comps <- Comps %>%
  filter(competition_name == "FIFA World Cup", competition_gender == "male", season_name %in% c("1970", "2022"))

# Get all matches for the FIFA World Cup
Matches <- FreeMatches(Comps)

# Download all event data for these matches
StatsBombData <- free_allevents(MatchesDF = Matches, Parallel = T)

# Clean the data
StatsBombData <- allclean(StatsBombData)

```

Join competition data to get seasons (year)

```{r}
# Filter only shots from StatsBombData
shots   <- StatsBombData %>% filter(type.name == "Shot")
passes  <- StatsBombData %>% filter(type.name == "Pass")

#Joined competetion to get seasons (for year analysis)

#Joined shots with competions, to obtain seasons
shots   <- shots %>% left_join(Comps, by = c("competition_id", "season_id"))

#Joined passes with competions, to obtain seasons
passes  <- passes %>%left_join(Comps, by = c("competition_id", "season_id"))
```

### Shot Density Map in the FIFA World Cup (1970 vs. 2022)

In this analysis, we group shots into a 10x10 meter grid and examine how shooting patterns have evolved in the FIFA World Cup (1970 vs. 2022).

```{r}
# Ensure shot locations are within pitch boundaries
heatmap <- shots %>%
  mutate(
    location.x = pmax(0, pmin(120, location.x)),
    location.y = pmax(0, pmin(80, location.y))
  ) %>%
  mutate(
    xbin = cut(location.x, breaks = seq(0, 120, by = 10), include.lowest = TRUE, labels = FALSE),
    ybin = cut(location.y, breaks = seq(0, 80, by = 10), include.lowest = TRUE, labels = FALSE)
  )

# Aggregate shots by bin and season
heatmap_summary <- heatmap %>%
  group_by(xbin, ybin, season_name) %>%
  summarise(
    shot_count = n(),
    location.x = (xbin - 1) * 10 + 5,  # Center each bin
    location.y = (ybin - 1) * 10 + 5,
    .groups = "drop"
  )

```

This heatmap below visualizes shot density across different FIFA World Cup seasons.

-   Shot locations are binned into a 10x10 meter grid.

-   The color scale (blue to red) represents the concentration of shots, with red areas indicating high shot frequency.

-   The facet wrap by season allows a direct comparison of shooting patterns across different tournaments.

```{r}

ggplot(data = heatmap_summary, aes(x = location.x, y = location.y, fill = shot_count)) +
  geom_tile(width = 10, height = 10, alpha = 0.9, color = "black") +  

  # Pitch
  annotate("rect", xmin = 0, xmax = 120, ymin = 0, ymax = 80, fill = NA, colour = "black", size = 0.6) +
  annotate("rect", xmin = 18, xmax = 0, ymin = 18, ymax = 62, fill = NA, colour = "black", size = 0.6) +
  annotate("rect", xmin = 102, xmax = 120, ymin = 18, ymax = 62, fill = NA, colour = "black", size = 0.6) +
  annotate("segment", x = 60, xend = 60, y = 0, yend = 80, colour = "black", size = 0.6) +
  annotate("point", x = 60, y = 40, colour = "black", size = 1.5) +
  annotate(
    "path",
    x = 60 + 10 * cos(seq(0, 2 * pi, length.out = 100)),
    y = 40 + 10 * sin(seq(0, 2 * pi, length.out = 100)),
    colour = "black", size = 0.6
  ) +

  scale_fill_gradient(low = "blue", high = "red") +
  scale_y_reverse() +
  labs(
    title = "Shot Density Map 1970-2022",
    subtitle = "All Shots – FIFA World Cup"
  ) +

  coord_fixed(ratio = 95 / 100) +

  facet_wrap(~season_name, ncol=1) +

  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "right",  # Move legend below plot
    plot.background = element_rect(fill = "white"),
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(10, 10, 10, 10),
    strip.text = element_text(size = 14, face = "bold")  # Bigger facet titles

  )
```

#### **Conclusions from the Shot Density Map (1970 vs. 2022)**

-   Shot Concentration Shift

    -   1970: Shots are more evenly distributed across different areas, including the midfield and wide positions.

    -   2022: A high concentration inside the penalty box, particularly in central positions (red zone), indicating a shift toward more efficient shot locations.

-   Increase in High-Value Shooting Areas

    -   The 2022 map has more intense red zones, showing higher shot frequency in dangerous areas near the goal.

    -   1970 features more long-range attempts, particularly near the center circle, whereas 2022 focuses on close-range efforts.

-   Tactical Evolution Over Time

    -   The 1970s relied more on long shots and wide attacks, suggesting a more open-play style.

    -   2022 aligns with modern tactics, prioritizing high-percentage finishing inside the box for better goal-scoring efficiency.

## Lionel Messi Shots in 2022 FIFA World

To analyze Lionel Messi's shooting performance in the 2022 FIFA World Cup, we first filter his shots from the dataset.

```{r}
library(SBpitch)

# Filter shots from Lionel Messi inside the box
shots.messi <- shots %>%
  filter(
    type.name == "Shot",
    player.id == 5503  # Lionel Messi's ID
  ) %>%
  mutate(
    goal = ifelse(shot.outcome.name == "Goal", "Goal", "Missed")  # Define goal/miss categories
  )

```

The following plot shows the **trajectories of Messi’s shots** during the tournament, with goals highlighted in red and missed shots in blue.

```{r}
# Plot pitch with colored shots
create_Pitch() +
  geom_segment(
    data = shots.messi, 
    aes(
      x = location.x, 
      y = location.y,
      xend = shot.end_location.x, 
      yend = shot.end_location.y,
      color = goal  # Color by goal/miss
    ),  
    lineend = "round", 
    size = 0.5,
    arrow = arrow(length = unit(0.07, "inches"), ends = "last", type = "open")
  ) +
  
  # Define colors for shot outcomes
  scale_color_manual(values = c("Goal" = "red", "Missed" = "blue")) +  

  # Labels and formatting
  labs(
    title = "Lionel Messi",
    subtitle = "FIFA World Cup Shots, 2002",
    color = "Shot Outcome"
  ) +

  # Adjust field proportions
  scale_y_reverse() +
  coord_fixed(ratio = 105 / 100)+
   theme(
    legend.position = "bottom",  # Places legend below
    legend.title = element_text(face = "bold"),  # Bold legend title
    legend.key.width = unit(2.5, "cm"),  # Adjust legend size
    legend.text = element_text(size = 10)  # Set legend text size
  )

```

**Conclusions from Lionel Messi Shots in 2022 FIFA World**

-   The majority of Messi’s shots originate from inside or near the penalty area.

-   There are no significant attempts from outside the box, meaning Messi relied more on close-range finishing rather than long-distance shots.

-   Goals (Red Arrows): Concentrated inside the penalty area, particularly near the center.

### **Brazil 1970 vs 2022: Shot Map Comparison**

The code below filters shot data for the Brazil national team in the 1970 and 2022 FIFA World Cups, excluding penalties. It also defines a color scale for Expected Goals (xG) to visually represent shot quality. The filtered dataset includes shot locations, xG values, body parts used, and season names for further analysis and visualization.

```{r}
# Set team of interest
selected.team <- "Brazil"

# Define Expected Goals color scale
shotmapxgcolors <- c("#192780", "#2a5d9f", "#40a7d0", "#87cdcf", "#e7f8e6", 
                     "#f4ef95", "#FDE960", "#FCDC5F", "#F5B94D", "#F0983E", 
                     "#ED8A37", "#E66424", "#D54F1B", "#DC2608", "#BF0000", 
                     "#7F0000", "#5F0000") 

# Filter shots for selected team (excluding penalties)
shots.brazil <- shots %>%
  filter((shot.type.name != "Penalty" | is.na(shot.type.name)) & 
         team.name == selected.team  & season_name %in% c("1970", "2022")) %>%
  select(location.x, location.y, shot.statsbomb_xg, shot.body_part.name, season_name)
```

This visualization compares Brazil's shot locations and expected goals (xG) in the 1970 and 2022 FIFA World Cups.

-   Shot locations are plotted with colors representing their xG value (likelihood of scoring).

-   Different shapes indicate the body part used (head, left foot, right foot).

-   Facet wrap by season allows a direct visual comparison between both eras.

This analysis highlights how Brazil's shot-taking patterns and quality evolved over time.

```{r}
# Plot
ggplot() +
  # Draw pitch
  annotate("rect", xmin = 0, xmax = 120, ymin = 0, ymax = 80, fill = NA, colour = "black", size = 0.6) +
  annotate("rect", xmin = 18, xmax = 0, ymin = 18, ymax = 62, fill = NA, colour = "black", size = 0.6) +
  annotate("rect", xmin = 102, xmax = 120, ymin = 18, ymax = 62, fill = NA, colour = "black", size = 0.6) +
  annotate("point", x = 108, y = 40, colour = "black", size = 1.05) +  # Penalty spot
  annotate("point", x = 60, y = 40, colour = "black", size = 1.05) +   # Center spot

  # Plot shots
  geom_point(data = shots.brazil, 
             aes(x = location.x, y = location.y, fill = shot.statsbomb_xg, shape = shot.body_part.name),
             size = 2, alpha = 0.8) +

  # Color and shape scales
  scale_fill_gradientn(colours = shotmapxgcolors, limits = c(0, 0.8), oob = scales::squish, name = "Expected Goals") +
  scale_shape_manual(values = c("Head" = 21, "Right Foot" = 23, "Left Foot" = 24), name = "Body Part") +

  # Facet by season
  facet_wrap(~ season_name) +

  # Labels and theme
  labs(title = paste(selected.team, "Shot Maps"),
       subtitle = "FIFA World Cup, Different Years") +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    aspect.ratio = 65 / 100
  ) +
  coord_flip(xlim = c(85, 125))

```

#### **Conclusions from Brazil 1970 vs 2022: Shot Map Comparison**

-   Brazil in 1970 relied more on long-range shooting and varied attempts.

-   In 2022, shot selection improved, favoring higher xG attempts inside the penalty box.

-   This shift aligns with modern football trends, emphasizing efficiency over volume.

### 
